---
title: "homework-3"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{homework-3}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(bis557)
devtools::load_all()
```

## Problem 1

Constructing a well-conditioned Hessian along with an ill-conditioned logistic Hessian.

```{r}
X <- c(-3.5, -3, -3.6, 6, 8)/2
X <- cbind(rep(1,5),X)
H <- t(X) %*% X
kappa(H)
```

```{r}
betas <- c(0.1,100)
X <- c(-3.5, -3, -3.6, 6, 8)/2
X <- cbind(rep(1,5),X)
mu <- 1 / (1+exp(-X %*% betas))
D <- diag(x=as.vector(mu), nrow=5, ncol=5)
H <- t(X) %*% D %*% X
kappa(H)
```

As seen, the D matrix inflates the condition number by more than 100 times.

## Problem 2

First generate some random data for testing.

```{r}
n <- 1000
p <- 3
betas <- c(0.2,2,1)
X <- cbind(1, matrix(rnorm(n*(p-1)), ncol=p-1))
mu <- 1 / (1 + exp(-X %*% betas))
y <- as.numeric(runif(n) > mu)
```
 
I first built a GLM fitting function using gradient descent (without hessian), optimizing log-likelihood. 
 
```{r}
fit <- GLMgradient(X=X,y=y,mu_fun=function(eta) 1/(1+exp(-eta)), T_fun=identity, lrate=0.01, maxiter=10000, tol=1e-4)
print(fit)
```
 
The fitting is reasonably good.

Next I use optimization with momentum.

```{r}
fit <- GLMgradientMomentum(X=X,y=y,mu_fun=function(eta) 1/(1+exp(-eta)), T_fun=identity, mom=0.2, lrate=0.01, maxiter=10000, tol=1e-4)
print(fit)
```
 
Here the fitting accuracy is a little better than the one with constant step size.

## Problem 3

I implemented a softmax regression to generalize logistic regression for classifying more than 2 classes.

First generate some random data for testing.

```{r}
library(ramify)  # call argmax function
n <- 1000
p <- 3
betas <- matrix(c(0.2,2,1,-1,0.2,0.5,0.5,1.0,-0.5), nrow=3, ncol=3)
X <- cbind(1, matrix(rnorm(n*(p-1)), ncol=p-1))
z <- exp(X %*% betas)
z <- z/colSums(z)
y <- argmax(z)
```

Then I fit data using the implemented softmaxreg function.

```{r}
fit <- softmaxreg(X=X,y=y,lrate=0.001, maxiter=10000, tol=1e-4)
print(fit)
```
 
 There is an issue.
 

